The Causal Effect of Media Frames on Issue Attitudes

### 2021-09-28 paper/main.tex

\pagestyle{fancy}
\fancyhf{}
\rhead{Can Meaning Travel?}
\lhead{Nicolai Berk}
\cfoot{\thepage}

@@

\title{Can Meaning Travel? The Causal Effect of Media Frames on Issue Definitions and Attitudes}
\author{Nicolai Berk\footnote{Doctoral Candidate, Dynamics Doctoral Program, Humboldt-Universität zu Berlin, \href{mailto:nicolai.berk@hu-berlin.de}{nicolai.berk@hu-berlin.de}}}
\date{September 2021}

@@

\begin{abstract}
    A large body of literature debates the existence and extent of effects of media framing on consumers' issue attitudes. At the same time, mechanisms of opinion formation have mainly been studied experimentally. Combining panel data from the German Longitudinal Election Study with fine-grained analyses of media framing, I address both debates with observational data. I provide evidence from 1,056 fixed-effect, difference-in-differences models suggesting systematic differences in newspaper readers' opinion shifts. Second, I show that such changes in consumers' attitudes do \textit{not} correspond to changes in the issue framing of the consumed media outlet. Instead, these changes could be explained by an ideological readership reacting differently to changes in the news agenda. Lastly, the study explores whether such changes in media framing affect consumers' issue definitions using state-of-the-art text analysis methods. The findings contribute to our understanding of opinion formation processes and the relevance of the news media in the 21st century.
\end{abstract}

@@

\begin{itemize}
  \item starting point: historical discussion on minimal vs maximal media effects and the discussion on external validity
  \item agenda setting yes (cite), but what about framing? Debate goes on
  \item second: unclear mechanisms - look for lab evidence to cite on this
\end{itemize}

@@

argument 1: key issue is not the external validity of framing experiments, but more likely to observe framing effects when consuming the SAME newspaper, which one trusts already - basically cue taking (Foos, Spirig, Ladd vs Gentzkow, Guess, resolved by Slothuus and Leeper)

\subsection{Mechanisms of framing}

argument 2: opinion shifts come about through changing associations (framing/issue definition) $\rightarrow$ value expectancy framework, but with a neurological basis

\subsection{Implications for the study of framing effects}

\begin{itemize}
\item Lesson: more observational data necessary, but we need to study cases where the slant of a newspaper that people already read changes, without obvious changes eg in ownership that made readers discount the slant
\item people could a) not react, b) stop reading their newspaper, c) change their opinion
\item changes in issue association (in the news) should affect people's issue definitions, which changes their opinion on the issue.
\end{itemize}

@@

\subsection{Data and case description}

\subsubsection{Public opinion data}

% plot with opinion development per paper for four / six issues

\begin{itemize}
    \item GLES Panel 2016-2020
    \item about 10,000 - 15,000 respondents per wave
    \item data on issue attitudes, importance, and news consumption
\end{itemize}

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/Immigration_papers.png}
    \caption{Opinions on migration across time (higher values indicate more restrictive attitudes).}
    \label{fig:issues}
\end{figure}

\subsubsection{Data on news framing}
% plot with frames and survey dates
\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/frames_papers_wtd_focus.png}
    \caption{Prevalence of different migration frames across newspapers. Gray vertical lines indicate survey field dates.}
    \label{fig:frames}
\end{figure}

@@

\subsubsection{Have readers' issue definitions changed as a result of the changing media frames?}

The final analysis in this paper aims to show that changes in readers' migration definitions correspond to changes in news coverage. This can be achieved by applying embedding regression to open survey responses, which shows differing word associations for subgroups of documents. This would show if respondents reading a certain newspaper define migration more in line with e.g. the humanitarian situation in refugee camps, compared to a preceding wave and controlling for shifts in other newspapers.

@@

\subsection{Correlation of DiD with frame prevalence}
% show correlation of DiD with prevalence of media frames (STM)
\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/effectplot_noimp.png}
    \caption{Correlation of opinion shifts (DiDs) with media attention or shifts in attention (DiD's) for different specifications.}
    \label{fig:did_corr}
\end{figure}

@@

The analysis shows that, generally, no clear pattern can be observed in the correlation of these effects. Independent of the choice of the dependent variable, specification of the independent variable (DiD/attention and differing periods of news coverage), and across all migration frames, t-values scatter around 0 (see also \ref{fig:t_values} in the appendix). This constitutes strong evidence against an effect of news framing on opinion formation. Even changes in issue salience in a given newspaper do not correspond to changes in readers' issue attitudes (see bottom two panels).

A second analysis will estimate the decrease in explanatory value of the wave-news consumption interaction/DiD when media attention is added as a control.

\subsection{Shifts in issue definitions}

% show for one specific case that shift in media frame corresponded to

\section{Discussion}

@@

\begin{figure}
    \centering
    \includegraphics[width = \textwidth]{paper/vis/t_values_noimp.png}
    \caption{Caption}
    \label{fig:t_values}
\end{figure}

>>>>

\pagestyle{fancy}
\fancyhf{}
\rhead{The Causal Effect of Media Frames}
\lhead{Nicolai Berk}
\cfoot{\thepage}

@@

\title{The Causal Effect of Media Frames on Issue Attitudes}
\author{Nicolai Berk\footnote{Doctoral Candidate, Dynamics Doctoral Program, Humboldt-Universität zu Berlin, \href{mailto:nicolai.berk@hu-berlin.de}{nicolai.berk@hu-berlin.de}}}
\date{September 2021}

@@

\begin{abstract}
    A large body of literature debates the existence and extent of effects of media framing on consumers' issue attitudes. Combining panel data from the German Longitudinal Election Study with a fine-grained analyses of media framing, I address this debate with observational data, providing evidence from 1,752 fixed-effect, difference-in-differences models suggesting systematic differences in newspaper readers' opinion shifts. Second, I show that such changes in consumers' attitudes towards immigration surrounding the German national election 2017 did \textit{not} correspond to changes in the migration framing of consumed media outlets. The findings contribute to our understanding of opinion formation processes and the relevance of the news media in the 21st century.
\end{abstract}

@@

How does news media content affect citizens' issue attitudes? Research on political behaviour has discussed this question for nearly a century. Classic theories of agenda setting hold that the attention devoted to issues in the news media will guide citizens to consider these issues important (\cite{McCombs1972}). Similarly, the literature on framing expects that the presentation of issues in the news media will affect recipients' opinions about these issues (\cite{Nelson1997}). Both theories suggest that media coverage strongly affects citizens' understanding of and opinions about politics.

Yet, a popular critique of these theories maintains that strong news effects are unlikely in the modern media environment, where consumers self-select into news sources of their liking. As a result, "most media users will rarely find themselves in the path of attitude-discrepant information". Even if consumers had to face such discrepant information, this should reinforce their existing views rather than persuade as they are used to a partisan media diet (\cite[724f]{Bennett2008}). Similarly, agenda setting effects are unlikely as a result of selective exposure (\cite{Lau2021}).

This conflicting evidence begs the question: when can we expect media framing effects? Some authors have argued that a lack of media effects can be explained by readers who discount a publications' bias. For example, when a left-wing newspaper endorses a left-wing candidate, this should have no effect on citizens' assessment of the candidate, as the endorsement was to be expected (\cite{Chiang2011a}, see also \cite{Spirig2020}). Additionally, citizens might take cues from their preferred publication because they developed trust in its coverage over time.

The present study contributes to this debate in two ways. First, instead of taking media slant as a given (\cite{Foos2020, Guess2021}), I measure the prevalence of different frames directly and assess whether their presence in the media across time affected readers' issue attitudes. Second, I use the variation in framing \textit{within} a given newspaper across time as the main independent variable. Many other studies explored effects of changes in newspaper consumption (\cite{Foos2020, Gentzkow2011}) or randomly assigned participants to consume certain news media rather than others (\cite{Guess2021}). However, readers usually do not change readership, nor are all newspapers equally credible to them. If trust in a given news source matters, or if readers' discount publications' bias, such designs are unlikely to measure how news consumption affects consumers' opinions. This paper hence assesses how the changing framing of a salient issue within an already consumed paper changes citizens' attitudes.

@@

A large amount of evidence underlines the importance of newspaper coverage for political opinion formation. Observational evidence shows that newspaper exposure (\cite{Foos2020, Spirig2020}), endorsements of specific candidates (\cite{Ladd2009a, Chiang2011a}), coverage of specific topics (\cite{King2017}), reports on political parties (\cite{Boomgaarden2009, Devine2020}), internet access (\cite{Schaub2020}), and changes in television news content (\cite{Durante2012}) have affected which issues citizens care about, their opinion about these issues, and their voting behaviour. An even larger body of experimental work reports strong effects of issue framing on participants' issue attitudes (see \cite{Busby2019} for a recent overview).

This "maximal effects" view is contrasted by a body of evidence indicating weak to absent media effects as well. \citeauthor{Gentzkow2011} show that even historically, the entry and exit of newspapers in the US does affect turnout, but that the newspapers' slant does not affect either party's vote share. They conclude that "the persuasive effect of partisan newspapers is limited" (\citeyear[3011]{Gentzkow2011}). Recently, \citeauthor{Guess2021} incentivised participants of an online-panel to visit left- or right-leaning news websites, with no effect on opinions (\citeyear{Guess2021}). An experiment by \citeauthor{Lau2021} indicates that increasing the likelihood of encountering information about a fictitious candidate's stance on a given issue does not increase the importance of that issue in reaching a voting decision, directly contradicting theories of agenda setting. This lack of evidence echos a general critique of the external validity of framing experiments and other experimental evidence about opinion formation (\cite{Barabas2010, Busby2019, Leeper2020}).

What might explain these different findings? Scope conditions for media effects seem to be surprisingly understudied given the contradicting body of evidence. \citeauthor{Chiang2011a} argue that endorsements from newspapers with ideological leanings neutral or opposite to the endorsed party/candidate should have stronger effects on voting behaviour, as voters discount the general bias of a newspaper (\citeyear{Chiang2011a}). \citeauthor{Spirig2020} shows that the takeover of a Swiss regional newspaper through a leading politician of the far-right and the subsequent change in its political slant had no effect on citizens' voting behaviour and suggests that this is a result of voters discounting the new owner's political slant and stopping to read the newspaper.

If citizens discount the content of newspapers, they should not react to news content which is randomly assigned to them. Instead, a most likely case for effects of news exposure would have to assess changing content \textit{within} the same newspapers and its effect on consumers' issue attitudes. This is precisely what this study aims to do.

\subsection{Agenda Setting}

\textbf{[Now that there seem to be no framing effects, I am considering to make this a broader study of media effects and include agenda setting. But I guess this might be hard to finish until the weekend, so I should probably drop it for now. Might still be relevant for future direction of the paper]}\medskip

[Insert discussion of McCombs \& Shaw, as well as recent studies on agenda setting through the media, might also contrast 'latent salience model' (similar to Gamson \& Modigliani) and 'direct agenda setting']

\medskip

\textbf{H1a (direct agenda setting):} The more an issue is emphasised in a given newspaper, the more important this issue is considered by readers of this newspaper, relative to other newspapers.

\textbf{H1b (latent salience model):} Increased issue attention in the news across time corresponds to increased importance among all readers, but higher attention in one newspaper compared to others does not increase the perceived importance among its readers relative to those of other newspapers (all move at the same time).

\subsection{Mechanisms of Framing}

While news slant is a very general concept, and sometimes operationalised as similarity to specific parties' communication (\cite{Gentzkow2010}) or issue positions/tonality (\cite{Spirig2020}), framing theory provides a specific expectation about which treatment should affect which issue attitudes how. As \citeauthor{Nelson1997} write: "framing is the process by which a communication source, such as a news organisation, defines and constructs a political issue or public controversy" (\citeyear[567]{Nelson1997}). A "frame" is then the operating unit of this process, defining the issue at hand in a specific way. Different forms of frames have been conceptualised. When I talk about 'frames' here, I refer to \textit{emphasis frames}. These frames emphasise certain topics in relation to an issue, guiding the recipient to think about the issue with those considerations in mind that are promoted by the frame (\cite[153f]{Leeper2020}). For example, issues of increasing welfare contributions might be discussed with reference to inequality and providing chances to the poor, or by mentioning that higher welfare benefits might result in higher taxes. Individuals' support of a statement will differ dependent on the frame presented (\cite{sniderman2004structure}).

While this broad definition gives a first idea of what framing is about, the concept of emphasis framing has been criticised for being too broad and imprecise to contribute anything meaningful to the field (\cite{Scheufele2012}). To address this shortcoming, I follow the framing literature building on the \textit{value-expectancy model} (\cite{Ajzen2000, Nelson1997}). This model suggests that an attitude on a given issue is a function of two things: considerations and their respective weights. The evaluation of the issue is subdivided into a number of considerations (to which political scientists might refer as "valence issues"), that will be easier to evaluate. For example, to decide whether a lockdown is necessary to battle the spread of Covid-19 in one's country, a person might consider the detrimental short-term effect on businesses and individuals' mental health (negative considerations), as well as the cost of overburdened hospitals and additional deaths (positive considerations, at least with respect to the evaluation of a lockdown). Individuals weigh each of these considerations to form an overall opinion on the subject.

Figure \ref{fig:cogStor} visualises this logic. A given issue is associated with an attitude or evaluation $y$. The issue is related to a number of different considerations, each of which carries an associated evaluation $x_i$, e.g. that the prevention of overcrowded hospitals is a good thing. Each of these considerations is more or less associated with the issue. Based on the strength of this association $w_i$, a given consideration's evaluation is more or less reflected in the issue evaluation $y$, which is equal to the weighted sum of considerations. Emphasis framing affects the issue attitude $y$ by changing the weights of different considerations\footnote{Note that this concept of framing is identical to second-level or attribute agenda setting (\cite{Lopezescobar2017, McCombs2000}; see \cite[174]{Mclaren2018} for a discussion of the overlap of these literatures). I will stick to the framing perspective here as the literature is focused on issues rather than candidates' attributes, and the conceptualisation as frame allows to form expectations about citizens' reactions to changing news frames.}.

This operationalisation allows us to generate specific expectations about how frames in the news media affect respondents' issue opinions. When presented with an emphasis frame on a given issue, the respondents' cognitive association of the issue with the emphasised consideration should increase, thus changing the evaluation of the issue by increasing the importance of the consideration. If the consideration is positive (towards the issue), the attitude towards the issue should improve, and the opposite when it is negative. \medskip

\textbf{H2 (emphasis framing):} If a newspaper emphasises more positive/negative frames of an issue, the readers' attitude about the issue should become more positive/negative.

% JB: I’m wondering if negative reference might help you with the definition. In our last colloquium you still got the criticism that framing was not well defined enough. I think you might be able to avoid a lot of the criticism (what makes framing different from issue ownership etc.) by addressing the contrast from framing from some of the more used concepts such as issue attention. Just an idea
% \footnote{I will stop here to outline the other chapters. However, a future version of this draft will discuss the cognitive basis of this model (considerations: affective storage as discussed by Taber and Lodge, Kunda; weights: physical cognitive association through Hebbian learning) and  whether this is a rational model (\cite{Ajzen2000, Druckman2001}).}.

\begin{center}
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{pres/vis/CognitiveStorage.png}
    \caption{The cognitive evaluation of an issue $y$ is determined by the strength of association $w_i$ with other concepts with existing evaluations $x_i$.}
    \label{fig:cogStor}
\end{figure}
\end{center}

@@

\subsection{Case description}

[describe context of German election 2017 and dominance of migration topic]

\subsection{Public opinion data}

To measure the dependent variable (issue attitudes) and differentiate the treatment groups (newspaper readers), I employ panel data from the German Longitudinal Election Study (GLES, \cite{GLES2019LongTermTracking}). The panel consists of 15 waves, of which 6 (4) contain variables on newspaper consumption and immigration (integration) attitudes. In each wave 9,000-15,000 respondents were interviewed, which allows the precise estimation of effects of media consumption.

Figure \ref{fig:issues} shows the two main dependent variables for each group of readers. Note that higher values on the immigration variable indicate more restrictive attitudes, while higher values on the integration variable are most liberal. As can be seen, readers of the tabloid 'Bild' differ strongly from consumers of other outlets, with far more restrictive attitudes. Both variables seem inversely correlated (as expected), with values moving towards the more liberal pole until autumn, where the trend is reversed. The development of issue importance across time can be found in Appendix \ref{app:importance}.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/issues_readers.png}
    \caption{Opinions on migration and integration across time (higher values on immigration indicate more restrictive attitudes, vice versa for integration).}
    \label{fig:issues}
\end{figure}

\subsection{Measuring news framing}

\subsubsection{Measuring news attention to migration}

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/salience_papers_focus.png}
    \caption{Share of migration articles out of all articles across time for different newspapers.}
    \label{fig:salience}
\end{figure}

\subsubsection{Measuring news frames}

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/frames_papers_focus.png}
    \caption{Prevalence of different migration frames across newspapers. Gray vertical lines indicate survey field dates.}
    \label{fig:frames}
\end{figure}

@@

@@

\subsection{Correlation of DiD with frame prevalence}
% show correlation of DiD with prevalence of media frames (STM)
\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/effectplot_frames.png}
    \caption{Correlation of opinion shifts (DiDs) with media attention or shifts in attention (DiD's) for different specifications.}
    \label{fig:did_corr}
\end{figure}

@@

% show for one specific case that shift in media frame corresponded to 
\begin{table}[!htbp] \centering 
  \caption{Individual-level Model} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{2}{c}{\textit{Dependent variable:}} \\ 
\cline{2-3} 
\\[-1.8ex] & Immigration attitude & Integration attitude \\ 
\\[-1.8ex] & (1) & (2)\\ 
\hline \\[-1.8ex] 
 Lagged DV & 0.793$^{***}$ & 0.664$^{***}$ \\ 
  &  & (0.004) \\ 
  & & \\ 
 Petty Crime & 0.0001 & 0.002 \\ 
  & (0.002) & (0.004) \\ 
  & & \\ 
 Capital Crime & 0.033$^{***}$ & $-$0.027$^{**}$ \\ 
  & (0.006) & (0.012) \\ 
  & & \\ 
 Refugee Numbers & $-$0.016$^{***}$ & $-$0.001 \\ 
  & (0.002) & (0.004) \\ 
  & & \\ 
 Refugee Camps & 0.025$^{***}$ & $-$0.004 \\ 
  & (0.004) & (0.012) \\ 
  & & \\ 
 Mediterranean S\&R & 0.035$^{***}$ & $-$0.005 \\ 
  & (0.002) & (0.005) \\ 
  & & \\ 
 Labour Market & 0.015$^{***}$ & 0.009 \\ 
  & (0.003) & (0.006) \\ 
  & & \\ 
 Deportation & $-$0.002$^{***}$ & 0.002$^{**}$ \\ 
  & (0.001) & (0.001) \\ 
  & & \\ 
 Constant & 0.045$^{***}$ & $-$0.288$^{***}$ \\ 
  & (0.015) & (0.036) \\ 
  & & \\ 
\hline \\[-1.8ex] 
Observations & 88,026 & 36,180 \\ 
R$^{2}$ & 0.617 & 0.421 \\ 
Adjusted R$^{2}$ & 0.617 & 0.421 \\ 
Residual Std. Error & 1.095 (df = 88017) & 1.095 (df = 36171) \\ 
F Statistic & 17,740.240$^{***}$ (df = 8; 88017) & 3,286.942$^{***}$ (df = 8; 36171) \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

\section{Discussion}
\begin{itemize}
    \item 
    \item people could a) not react, b) stop reading their newspaper, c) change their opinion
\end{itemize}

@@

\subsection{Descriptives on issue importance across time and newspaper consumption}\label{app:importance}

[add description and plot issue importance]
